---
title: "Reproducible Research: Peer assessment 2 - 
Impact of Severe Weather Events in the US Between 1950 and 2011"
output: 
  html_document:
    keep_md: true
---

## Sypnosis

### Background

The aim of this project is to explore the NOAA Storm Database and identify the types of events which are most damaging to population health and the economy.

This helps stakeholders such as the government, municipal manager etc to prioritise their resources allocated towards these events. 

### Data source

Data can be downloaded from the NOAA Storm Database at: <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>

More information regarding the database can be found here: <http://www.ncdc.noaa.gov/stormevents/choosedates.jsp?statefips=-999%2CALL>


## Data Processing

Firstly, load all the packages we need:
```{r,warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)

```

Secondly, read the data into a data frame.
```{r, cache=TRUE}

stormDataRaw <- read.csv(bzfile(file.path(getwd(),"repdata-data-StormData.csv.bz2")),
                         stringsAsFactors=FALSE)

stormDataRaw$EVTYPE <- as.factor(stormDataRaw$EVTYPE)
stormData <- tbl_df(stormDataRaw)
str(stormData)
```

Next, we need convert scalers into a numerical format (eg."k" represents 1,000). We use the unique function to look at the list of scalers we need to handle in relation to PROPDMG

```{r}
unique(stormData$PROPDMGEXP)
```

Based on this list of scalers, we do the numeric conversion accordingly.

```{r}
stormData[stormData$PROPDMGEXP=="H"|stormData$PROPDMGEXP=="h","PROPDMGEXP"] <- 100
stormData[stormData$PROPDMGEXP=="K","PROPDMGEXP"] <- 1000
stormData[stormData$PROPDMGEXP=="M"|stormData$PROPDMGEXP=="m","PROPDMGEXP"] <- 1000000
stormData[stormData$PROPDMGEXP=="B","PROPDMGEXP"] <- 1000000000
stormData[stormData$PROPDMGEXP=="0","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="1","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="2","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="3","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="4","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="5","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="6","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="7","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="8","PROPDMGEXP"] <- 10
stormData[stormData$PROPDMGEXP=="","PROPDMGEXP"] <- 1
stormData[stormData$PROPDMGEXP=="+","PROPDMGEXP"] <- 1
stormData[stormData$PROPDMGEXP=="-","PROPDMGEXP"] <- 1
stormData[stormData$PROPDMGEXP=="?","PROPDMGEXP"] <- 1

stormData$PROPDMGEXP <- as.numeric(stormData$PROPDMGEXP)
stormData <- mutate(stormData,property_damage=PROPDMG*PROPDMGEXP)
```

We repeat the steps to handle the scalers in relation to CROPDMG. First look at the list of scalers to determine what we have to convert.

```{r}
unique(stormData$CROPDMGEXP)
```

Based on this list of scalers, we do the numeric conversion accordingly.

```{r}
stormData[stormData$CROPDMGEXP=="K"|stormData$CROPDMGEXP=="k","CROPDMGEXP"] <- 1000
stormData[stormData$CROPDMGEXP=="M"|stormData$CROPDMGEXP=="m","CROPDMGEXP"] <- 1000000
stormData[stormData$CROPDMGEXP=="B","CROPDMGEXP"] <- 1000000000
stormData[stormData$CROPDMGEXP=="0","CROPDMGEXP"] <- 10
stormData[stormData$CROPDMGEXP=="2","CROPDMGEXP"] <- 10
stormData[stormData$CROPDMGEXP=="","CROPDMGEXP"] <- 1
stormData[stormData$CROPDMGEXP=="?","CROPDMGEXP"] <- 1

stormData$CROPDMGEXP <- as.numeric(stormData$CROPDMGEXP)
stormData <- mutate(stormData,crop_damage=CROPDMG*CROPDMGEXP)

```


## Results

### 1. Across the United States, which types of events are most harmful with respect to population health? 

Our objective is to see the fatalities and injuries caused by each weather event. 

```{r}
healthDmgTable <- group_by(stormData,EVTYPE) %>%
                  summarise(fatalities=sum(FATALITIES),injuries=sum(INJURIES))%>%
                  arrange(desc(fatalities+injuries))%>%
                  mutate(Total=fatalities+injuries)
head(healthDmgTable,10)
```
From the table above, we observe that tornadoes historically cause the greatest harm to population health in terms of fatalities and injuries.

To take a step further, we want to represent the fatalities and injuries caused by each weather event visually in the form of a stacked bar chart. However, due to space constraints we will only plot the top 10 events.  

To do that we have to convert the data to long format using the melt function from the reshape2 package:

```{r}
healthDmgTableLong <- melt(healthDmgTable[1:10,1:3],id.vars="EVTYPE")
healthDmgTableLong
```

Once we have the data in the right format, we are ready to plot the bar chart:

```{r,fig.width=9,fig.height=7}
healthDamage <- ggplot(data=healthDmgTableLong,
                        aes(x=reorder(EVTYPE,-value),y=value,fill=variable,order=-value))+
              geom_bar(stat="identity",position="stack",colour="black")+
              labs(x="Weather Event",y="Fatalities/Injuries Count",
                   title="Harm to population health caused by 
                          weather events in the US between 1950 and 2011")+
              theme(axis.text.x=element_text(angle=45,hjust=1),
                    plot.title=element_text(face="bold",vjust=1))+
              scale_y_continuous(breaks=seq(0,100000,by=5000))

healthDamage
# Note: The plot shows the top 10 weather events which cause the most harm to population health in terms of the number of fatalities and injuries caused.
```

Visually, we can observe that tornadoes cause the greatest harm to population health in terms of the number of fatalities and injuries caused. Besides tornadoes, other harmful weather events are excessive heat, tstm wind and so fourth.

### 2. Across the United States, which types of events have the greatest economic consequences?

Our objective is to see the property and crop damage caused by each weather event.

```{r}
economicDmgTable <- group_by(stormData,EVTYPE)%>%
                    summarise(TotalPropertyDamage=sum(property_damage)/10^9,
                              TotalCropDamage=sum(crop_damage)/10^9)%>%
                    arrange(desc(TotalPropertyDamage+TotalCropDamage))%>%
                    mutate(TotalDamage=TotalPropertyDamage+TotalCropDamage)

economicDmgTable
# The numbers shown are in US$ billions.
```
From the table above, we observe that floods have the greatest economic consequences in terms of property and crop damage.

To take a step further, we want to represent the property and crop damage caused by each weather event visually in the form of a stacked bar chart. However, due to space constraints we will only plot the top 10 events. 

To do that we have to convert the data to long format using the melt function from the reshape2 package:

```{r}
economicDmgTableLong <- melt(economicDmgTable[1:10,1:3],id.vars="EVTYPE")
economicDmgTableLong
```

Once the data is in the long format, we can plot our stacked bar chart:

```{r,fig.width=9,fig.height=7}
economicDmg <- ggplot(data=economicDmgTableLong,
                      aes(x=reorder(EVTYPE,-value),y=value,fill=variable))+
                geom_bar(stat="identity",position="stack",colour="black")+
                labs(x="Weather Event",y="Economic Damage (US$ billions)",
                title="Economic damage caused by weather events in the US 
                        between 1950 and 2011")+
                theme(axis.text.x=element_text(angle=45,hjust=1),
                      plot.title=element_text(face="bold"))+
                scale_fill_discrete(guide=guide_legend(reverse=TRUE))+
                scale_y_continuous(breaks=seq(0,170,by=10))
economicDmg
# Note: The plot shows the top 10 weather events which cause the most economic damage in terms of property and crop damage (in US$ billions)
```

Visually we can observe that floods historically cause the greatest economic damage in monetary terms. Besides floods, other economically damaging events are hurricanes/typhoons, tornadoes and so fourth.

