---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r,warning=FALSE,message=FALSE}
library(dplyr)
```

```{r}

if (!file.exists(file.path(getwd(),"activity.csv"))){
  unzip(file.path(getwd(),"activity.zip"),
        exdir=file.path(getwd()))
}

activityDataRaw <- tbl_df(read.csv(file.path(getwd(),"activity.csv"),stringsAsFactors=FALSE))
activityData <- filter(activityDataRaw,steps!="NA")
activityData$date <- as.Date(activityData$date,"%Y-%m-%d")

activityData
```

## What is mean total number of steps taken per day?

To calculate the total number of steps taken per day:
```{r}
totalStepsTakenPerDay <- group_by(activityData,date) %>%
                    summarise(total_steps=sum(steps))
totalStepsTakenPerDay
```


To make a histogram of the total number of steps taken each day: 
```{r}
hist(totalStepsTakenPerDay$total_steps,xlab="",main="Total number of steps taken each day",col="blue")
```


Mean of the total number of steps taken per day:
```{r}
mean(totalStepsTakenPerDay$total_steps)
```


Median of the total number of steps taken per day:
```{r}
median(totalStepsTakenPerDay$total_steps)
```


## What is the average daily activity pattern?

The time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r}
dailyActivityPatternTable <- group_by(activityData,interval)%>%
                              summarise(average_steps=mean(steps))

plot(x=dailyActivityPatternTable$interval,y=dailyActivityPatternTable$average_steps,
     type="l", main="Average Daily Activity Pattern",
     xlab="5 minute interval",
     ylab="Average number of steps taken, averaged across all days",col="blue")
```


The 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps:
```{r}
dailyActivityPatternTable[which.max(dailyActivityPatternTable$average_steps),]
```


## Imputing missing values
Total number of missing values in the dataset
```{r}
nrow(activityDataRaw[is.na(activityDataRaw$steps),])
```

Create new dataset by filling missing values with the average number of steps (across all days), for that 5-minute interval:
```{r}
activityDataFilled <- tbl_df(merge(x=activityDataRaw,
                            y=dailyActivityPatternTable,
                            by.x="interval",by.y="interval"))

activityDataFilled$steps[which(is.na(activityDataFilled$steps))]=
  activityDataFilled$average_steps[which(is.na(activityDataFilled$steps))]

activityDataFilled$date <- as.Date(activityDataFilled$date,"%Y-%m-%d")

activityDataFilled
```


Histogram of the total number of steps taken each day:
```{r}
totalStepsTakenPerDayFilled <- tbl_df(activityDataFilled) %>%
                                group_by(date) %>%
                                summarise(total_steps=sum(steps))

hist(totalStepsTakenPerDayFilled$total_steps,xlab="",main="Total number of steps taken each day (inputted missing data)",col="blue")
```


Mean total number of steps taken per day:
```{r}
mean(totalStepsTakenPerDayFilled$total_steps)
```


Median total number of steps taken per day:
```{r}
median(totalStepsTakenPerDayFilled$total_steps)
```

There is no impact of inputing missing data on the mean. However, inputing missing data increases the median.


## Are there differences in activity patterns between weekdays and weekends?

Create ```Weekday``` and ```Weekend``` factor variables in the dataset
```{r}
weekTable <- data.frame(Weekday=c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),
           WeekEndOrDay=c("Weekday","Weekday","Weekday","Weekday","Weekday","Weekend","Weekend"))

activityDataFilledWithWeekInfo <- tbl_df(activityDataFilled) %>%
                                    mutate(Weekday=weekdays(date))

dailyActivityPatternWithWeekInfo <- tbl_df(merge(x=activityDataFilledWithWeekInfo,
                                                 y=weekTable,by="Weekday")) %>%
                                    group_by(interval,WeekEndOrDay) %>%
                                    summarise(average_steps=mean(steps))

dailyActivityPatternWithWeekInfo
```


Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):
```{r}
library(lattice)
xyplot(average_steps~interval|WeekEndOrDay,data=dailyActivityPatternWithWeekInfo,
            layout=c(1,2),type="l",xlab="Interval",ylab="Number of steps")
```
