
#Code Book

This code book describes:

* Information about the raw data set
* Key variables of the data set
* Data cleaning operations done to clean the data

 
## Data Set Information:

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

A video of the experiment including an example of the 6 recorded activities with one of the participants can be seen in the following link: https://www.youtube.com/watch?v=XOEN9W05_4A

**Citation**:
Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. A Public Domain Dataset for Human Activity Recognition Using Smartphones. 21th European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

1. Smartlab - Non-Linear Complex Systems Laboratory
DITEN - Università degli Studi di Genova, Genoa (I-16145), Italy. 
2. CETpD - Technical Research Centre for Dependency Care and Autonomous Living
Universitat Politècnica de Catalunya (BarcelonaTech). Vilanova i la Geltrú (08800), Spain
activityrecognition '@' smartlab.ws

## Data Set Structure

|         Features      |         Subject             |           Activity                |
|-----------------------|-----------------------------|-----------------------------------|
|X_train.txt            |subject_train.txt            |y_train.txt                        |
|X_test.txt             |subject_test.txt             |y_test.txt                         |


## Key Variables

|       Key variables   |                           Description                       |
|-----------------------|-----------------------------------------------------------------|
|Subject                |ID from 1-30 denoting the person who took part in the experiment |
|Activity               |6 activities performed by the subject in file *activity_labels.txt*| 
|Features               |Measurements (mean,standard deviation,max,min etc.) of signals (tBodyAcc-XYZ, tGravityAcc-XYZ, tBodyAccJerk-XYZ etc). More information can be found in *features_info.txt*                                                                       |


## Data Cleaning Operations

#### 1. Merge the features, subject and activity tables into one data set ```data```
```{r eval=TRUE}
str(data)

```


#### 2. Create ```data_subset```, which is a subset of ```data``` containing only mean and standard deviation measurements. Modify the column names and ```data_subset$activity``` to make them more descriptive.
```{r eval=TRUE}
str(data_subset)
```

#### 3. Create an independent tidy data set which displays the average of each measurement for each ```subject``` and ```activity```.
```{r eval=TRUE}
str(data_tidy)

#Display the first 6 rows of the tidy data set
head(data_tidy,6)
```

